---
title: "Class 12 Lab"
author: "Alexis Galano (PID: A17628362)"
format: pdf
editor: visual
---

# Section 1. Proportion of G/G in a population

Here we read this CSV file

```{r}
mxl <- read.csv("373531-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
head(mxl)
```

```{r}
table(mxl$Genotype..forward.strand.) / nrow(mxl) * 100
```

```{r}
gbr <- read.csv("373522-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
head(gbr)
```

```{r}
round(table(gbr$Genotype..forward.strand.) / nrow(gbr) * 100, 2) 
```
This variant that is associated with childhood asthma is more frequent in the GBR population than the MKL population.

Lets now dig into this further.

## Section 4: Population Scale Analysis

How many samples do we have?

```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt")
head(expr)
```
```{r}
nrow(expr)
```
```{r}
table(expr$geno)
```
```{r}
summary(expr$exp)
```

- A13. The sample size for genotype A|A is 108, A|G is 233, and G|G is 121. The median expression level for the genotypes is 25.11.


```{r}
library(ggplot2)
```

Let's make a boxplot

```{r}
ggplot(expr) +
  aes(x=geno, y=exp, fill=geno) +
  geom_boxplot(notch = TRUE) +
  labs(x="Genotype", y="Expression")
```

- A14. According to the boxplot, the mean expression distributions of each genotype do not overlap signifiying that each genotype's expression value is different. Particularly, having a G|G genotype in this location is associated with having a lower ORMDL3 expression, while an A|A genotype is associated with a relatively higher ORMDL3 expression. Therefore, the SNP does effect the expression of the ORMDL3 gene.




