---
title: "Class 18: Pertussis and the CMI-PB project"
author: "Alexis Galano (PID: A17628362)"
format: pdf
editor: visual
---

# 1. Investigating pertussis cases by year

First we will examine and explore Perstussis case numbers in the US:

We can use the datapasta package to scrape this data from the website into R:

```{r}
#install.packages("datapasta")
```
> Q1. With the help of the R “addin” package datapasta assign the CDC pertussis case number data to a data frame called cdc and use ggplot to make a plot of cases numbers over time.

```{r}
cdc <- data.frame(
                                                     year = c(1922L,1923L,
                                                              1924L,1925L,1926L,
                                                              1927L,1928L,1929L,
                                                              1930L,1931L,1932L,
                                                              1933L,1934L,
                                                              1935L,1936L,1937L,
                                                              1938L,1939L,1940L,
                                                              1941L,1942L,1943L,
                                                              1944L,1945L,1946L,
                                                              1947L,1948L,1949L,
                                                              1950L,1951L,1952L,
                                                              1953L,1954L,
                                                              1955L,1956L,1957L,
                                                              1958L,1959L,1960L,
                                                              1961L,1962L,1963L,
                                                              1964L,1965L,1966L,
                                                              1967L,1968L,1969L,
                                                              1970L,1971L,1972L,
                                                              1973L,1974L,
                                                              1975L,1976L,1977L,
                                                              1978L,1979L,1980L,
                                                              1981L,1982L,1983L,
                                                              1984L,1985L,1986L,
                                                              1987L,1988L,1989L,
                                                              1990L,1991L,1992L,
                                                              1993L,1994L,
                                                              1995L,1996L,1997L,
                                                              1998L,1999L,2000L,
                                                              2001L,2002L,2003L,
                                                              2004L,2005L,2006L,
                                                              2007L,2008L,2009L,
                                                              2010L,2011L,2012L,
                                                              2013L,2014L,
                                                              2015L,2016L,2017L,
                                                              2018L,2019L,2020L,
                                                              2021L),
                             cases = c(107473,164191,
                                                              165418,152003,
                                                              202210,181411,161799,
                                                              197371,166914,
                                                              172559,215343,179135,
                                                              265269,180518,
                                                              147237,214652,227319,
                                                              103188,183866,
                                                              222202,191383,191890,
                                                              109873,133792,
                                                              109860,156517,74715,
                                                              69479,120718,68687,
                                                              45030,37129,
                                                              60886,62786,31732,
                                                              28295,32148,40005,
                                                              14809,11468,17749,
                                                              17135,13005,6799,
                                                              7717,9718,4810,
                                                              3285,4249,3036,3287,
                                                              1759,2402,1738,
                                                              1010,2177,2063,
                                                              1623,1730,1248,1895,
                                                              2463,2276,3589,
                                                              4195,2823,3450,
                                                              4157,4570,2719,4083,
                                                              6586,4617,5137,
                                                              7796,6564,7405,
                                                              7298,7867,7580,9771,
                                                              11647,25827,25616,
                                                              15632,10454,
                                                              13278,16858,27550,
                                                              18719,48277,28639,
                                                              32971,20762,17972,
                                                              18975,15609,18617,
                                                              6124,2116)
                           )



```


```{r}
head(cdc)
```
I want a plot of cases per year with ggplot

```{r}
library(ggplot2)
```


> Q2. Using the ggplot geom_vline() function add lines to your previous plot for the 1946 introduction of the wP vaccine and the 1996 switch to aP vaccine (see example in the hint below). What do you notice?

```{r}
ggplot(cdc) + 
  aes(year, cases) +
  geom_line() +
  geom_point() +
  geom_vline(xintercept = 1947, color= "blue") +
  geom_vline(xintercept = 1992, color= "red") +
  labs(title="Pertussis Cases by Year (1922-")
  
```

- A2. I notice that there is a decrease in cases from 1947 to 1992.


> Q3. Describe what happened after the introduction of the aP vaccine? Do you have a possible explanation for the observed trend?

- A3. After the introduction of the aP vaccine in 1992 the trend increases slightly. This may be due to individuals not wanting to try the new vaccine because of doubt or fear.


Access data from the CMI-PB project 

This database (like many modern projects) uses an API to return JSON format data.

We will use the R package called 'jsonlite'
```{r}
library(jsonlite)

subject <- read_json("http://cmi-pb.org/api/subject", simplifyVector = TRUE)

head(subject)
```

Q. How many wP (older) individuals and aP (newer) individuals are in this dataset?
> Q4. How many aP and wP infancy vaccinated subjects are in the dataset?

```{r}
subject$infancy_vac
```

```{r}
table(subject$infancy_vac)
```
- A4. There are 118 total infancy vaccinated subjects, 60 aP and 58 wP.



> Q5. How many Male and Female subjects/patients are in the dataset?

```{r}
table(subject$biological_sex)
```
- A5. There are 79 female and 39 male subjects in the dataset.


Q. What is the number of individuals by biological sex and race?

> Q6. What is the breakdown of race and biological sex (e.g. number of Asian females, White males etc…)?

```{r}
table(subject$race, subject$biological_sex)
```
- A6. The table above shows the breakdown of race and biological sex among the subject population.

```{r}
subject$year_of_birth
```
```{r}
#install.packages("lubridate")
```

```{r}
library("lubridate")
```

We can use the lubridate package to ease the pain of doing math with dates

```{r}
today()
```
```{r}
# year, month, day format
today() - ymd("2000-01-01")
```
```{r}
# month, day, year format
today() - mdy("12-12-2001")
```

```{r}
time_length( today() - ymd("2001-12-12"),  "years")
```

So what is the age of everyone in the dataset?
> Q8. Determine the age of all individuals at time of boost?

```{r}
subject$age <- time_length(today() - ymd(subject$year_of_birth), "years")
head(subject$age)
```
A8. See head() below
```{r}
int <- ymd(subject$date_of_boost) - ymd(subject$year_of_birth)
age_at_boost <- time_length(int, "year")
head(age_at_boost)
```

> Q9. With the help of a faceted boxplot or histogram (see below), do you think these two groups are significantly different?

```{r}
ggplot(subject) + 
  aes(age, fill=infancy_vac) + 
  geom_histogram(show.legend = FALSE) + 
  facet_wrap(vars(infancy_vac), nrow = 2) + 
  xlab("Age in years")
```
- A9. Yes it appears that these two groups are significantly different because they are defined histograms.

## Get more data from CMI-PB

```{r}
specimen <- read_json("https://www.cmi-pb.org/api/specimen", simplifyVector = TRUE) 
head(specimen)
titer <- read_json("https://www.cmi-pb.org/api/specimen", simplifyVector = TRUE) 
```


We need to **join** these two tables (subject and specimen) to make a single new "meta" table with all our metadata. We will use the 'dplyr' join functions to do this

```{r}
library("dplyr")

# inner_join combines 2 data with only categories that we need)
meta <- inner_join(subject, specimen)
dim(meta)
head(meta)
```


```{r}
ab_titer <- read_json("https://www.cmi-pb.org/api/v4/plasma_ab_titer", simplifyVector = TRUE)
head(ab_titer)
```

> Q10. Now using the same procedure join meta with titer data so we can further analyze this data in terms of time of visit aP/wP, male/female etc.

One more 'inner_join()' to add all our metadata in 'meta' on to our 'ab_data' table:

```{r}
abdata <- inner_join(ab_titer, meta)
head(abdata)
```
- A10. See head() above.

```{r}
igg <- abdata %>% filter(isotype == "IgG")
head(igg)
```

> Q11. How many specimens (i.e. entries in abdata) do we have for each isotype?

```{r}
table(abdata$isotype)
```
- A11. See table() above.

Our first exploratory plot:

```{r}
table(abdata$antigen)
```

> Q13. Complete the following code to make a summary boxplot of Ab titer levels (MFI) for all antigens:

```{r}
ggplot(igg) +
  aes(MFI, antigen, col = infancy_vac) +
  geom_boxplot() +
    xlim(0,75) +
  facet_wrap(vars(visit), nrow=2)
```

Why are certain antigens and not others very variable in their detected levels here?

Can you facet or even just color by infancy_vac? Is there some difference?

There are potentially some differences here but in general it is hard to tell with this whole dataset overview...


```{r}
table(abdata$dataset)
```
Let's focus in on just the 2021_dataset.

```{r}
abdata.21 <- filter(abdata, dataset == "2021_dataset")
table(abdata.21$dataset)
```

Focus on PT antigen IgG levels
```{r}
pt.21 <- filter(abdata.21, isotype == "IgG",  antigen == "PT")
```


Plot of days (time) relative to boost vs MFI levels
```{r}
ggplot(pt.21) + 
  aes(x=planned_day_relative_to_boost,
        y=MFI_normalised,
        col=infancy_vac,
        group=subject_id) +
  geom_point() +
  geom_line()
```

```{r}
abdata.21 <- abdata %>% filter(dataset == "2021_dataset")

abdata.21 %>% 
  filter(isotype == "IgG",  antigen == "PT") %>%
  ggplot() +
    aes(x=planned_day_relative_to_boost,
        y=MFI_normalised,
        col=infancy_vac,
        group=subject_id) +
    geom_point() +
    geom_line() +
    geom_vline(xintercept=0, linetype="dashed") +
    geom_vline(xintercept=14, linetype="dashed") +
  labs(title="2021 dataset IgG PT",
       subtitle = "Dashed lines indicate day 0 (pre-boost) and 14 (apparent peak levels)")
```


> Q17. Do you see any clear difference in aP vs. wP responses?

- A17. Yes based on the ggplot, we can see great differences in aP vs wP responses. wP infancy_vac has a higher normalized MFI compared to aP infancy_vac. At the dashed line there is higher wP normalized MFI peak compared to aP.








